mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-01 Draft1.Rmd
operators

@@

@@

* General introduction on R and its features
* What is mathml 
* issues the user must be aware of
* points of strength
* future directions
* (inspirational ending)
* ways to install rolog and mathml

>>>>

## Operators
A list of possible operators is available here.

```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}

catmath(sum(x[i], i==1L, n))
catmathjax(sum(x[i], i==1L, n))

# fixme: braces not needed when right to *
catmath(a * sum(b))
catmathjax(a * sum(b))

# Multiplication (capital Pi) is missing

# Integral is missing

catmath(A & B)
catmathjax(A & B)

catmath(A | B)
catmathjax(A | B)

catmath(!A)
catmathjax(!A)

catmath(xor(A, B))
catmathjax(xor(A, B))

catmath(exp(2L*pi*i))
catmathjax(exp(2L*pi*i))

catmath(expm1(2L*pi*i))
catmathjax(expm1(2L*pi*i))

catmath(log(x))
catmathjax(log(x))

catmath(log10(x))
catmathjax(log10(x))

catmath(log2(x))
catmathjax(log2(x))

catmath(logb(x, e))
catmathjax(logb(x, e))

catmath(log1p(x))
catmathjax(log1p(x))

catmath(ceiling(pi))
catmathjax(ceiling(pi))

catmath(floor(pi))
catmathjax(floor(pi))

catmath(sin(pi/2L))
catmathjax(sin(pi/2L))

catmath(cos(pi/2L))
catmathjax(cos(pi/2L))

catmath(tan(pi/2L))
catmathjax(tan(pi/2L))

## Absolute value

Is it also possible to calculate and represent the absolute value, as:

```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
catmath(abs(x))
catmathjax(abs(alpha))

@@

## Mistakes

The case study presented above displays several possible ways of how \pkg{mathml} can be a useful tool in data representation and calculation. Furthermore, the careful reader might have also noticed that \pkg{mathml} allows also for a more smoothly detection of possible mistakes in writing.

An interesting use case of the package is in this sense the readability of self-written
functions in R code. R code is notoriously hard to read, as are most programming
languages [@green1977]. The poor legibility of the language is one of the main
sources of mistakes. For illustration, we consider Equation 10 in
Schwarz [-@schwarz1994], namely the same equation reported above. The first version has a wrong parenthesis, which is barely visible in the R code.

````{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
f <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
}

attributes(f)$name = quote(f)
mathml(f)
````

But, if the function is rendered in MathML, the wrong parenthesis appears as a curly
brace, and we immediately see the mistake. The correct version is shown below
for comparison.

````{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
f <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
}

attributes(f)$name = quote(f)
mathml(f)
````

This example may seem trivial. However, as the reader well knows, a missed parenthesis is often the cause of untraceable errors and contradictory results. Thus, rendering the expression in MathMl improves the readability of the text, allowing better identification of the sources of possible errors.

@@

This package expands the possibilities of R\ by allowing calculations and graphical rendition at the same time. This makes it easier for the user to notice writing errors and correct them more smoothly. Building on current features of R\ and existing packages for displaying formulas in R\ (e.g. Murrell and Ihaka [-@murrell2000], Allaire et al. [-@allaire2018]), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility. To do so, it uses the Prolog's grammar rules, via the \pkg{rolog}, and translates the R\ expressions to MathML. Then, two possible solutions for typesetting are available: LaTeX and MathJax to export R\ Markdown documents in different formats or ShinyApp webpages. The article and Appendix show examples of how the package can be used and what operations are currently possible. Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.
In its current version \pkg{mathml} presents some limitations, as, for instance, (...). However, several solutions to this issue are possible (...).

\pkg{mathml} is available for R\ version (…) and later, and can be easily installed using the usual `install.packages("mathml")`. The source code of the package is found at (…) including installation instructions for Unix, Windows and macOS.
